---
title: "Final Project Writeup"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



## 1. Exploratory data analysis (20 points)

We first performed basic checks to identify fields with (truly) missing data and potential abnormalities. The variable lot frontage was identified to have 282 or 18.8% missing entries. We removed this predictor, and verified at a latern stage that adding it in would not lead to better model predictions.

```{r read-data, echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)

load("ames_train.Rdata")

#print(paste0("The dataset has ", dim(ames_train)[1], " number of observations and ", dim(ames_train)[2], " features"))

#Variables with NA's and their proportion of missing data
miss = apply(is.na(ames_train), 2, sum)
miss_prop = round(miss[miss>0]/nrow(ames_train), 3)
#print(miss_prop) 
#which(miss_prop>0.5) # four features have greater than 50% of data "missing" -- drop these variables
```


```{r, echo=FALSE}
# Did not remove any NA entries in Lot.frontage

data=ames_train

data <- data  %>%
  #filter(!is.na(Lot.Frontage)) %>% 
  mutate(MS.SubClass= factor(MS.SubClass)) %>%
  mutate(Alley =  factor(Alley, levels = levels(addNA(Alley)), labels = c(levels(Alley), "None"), exclude = NULL)) %>%
  mutate(HouseAge = Yr.Sold- pmax(Year.Built, Year.Remod.Add))  %>%
  filter(!is.na(Mas.Vnr.Area))  %>% 
  mutate(Bsmt.YN = 1*(!is.na(Bsmt.Qual))) %>%
  mutate(Bsmt.Qual =  factor(Bsmt.Qual, levels = levels(addNA(Bsmt.Qual)), labels = c(levels(Bsmt.Qual), "None"), exclude = NULL)) %>%
  mutate(Bsmt.Qual = relevel(Bsmt.Qual, ref="None")) %>%
  mutate(Bsmt.Cond =  factor(Bsmt.Cond, levels = levels(addNA(Bsmt.Cond)), labels = c(levels(Bsmt.Cond), "None"), exclude = NULL)) %>%
  mutate(Bsmt.Cond = relevel(Bsmt.Cond, ref="None")) %>%
  mutate(Bsmt.Exposure =  factor(Bsmt.Exposure, levels = levels(addNA(Bsmt.Exposure)), labels = c(levels(Bsmt.Exposure), "None"), exclude = NULL)) %>%
  mutate(Bsmt.Exposure = relevel(Bsmt.Exposure, ref="None")) %>%
  mutate(BsmtFin.Type.1=  factor(BsmtFin.Type.1, levels = levels(addNA(BsmtFin.Type.1)), labels = c(levels(BsmtFin.Type.1), "None"), exclude = NULL)) %>%
  mutate(BsmtFin.Type.1 = relevel(BsmtFin.Type.1, ref="None")) %>%
  mutate(BsmtFin.Type.2=  factor(BsmtFin.Type.2, levels = levels(addNA(BsmtFin.Type.2)), labels = c(levels(BsmtFin.Type.2), "None"), exclude = NULL)) %>%
  mutate(BsmtFin.Type.2 = relevel(BsmtFin.Type.2, ref="None")) %>%
  mutate(X12.SF= X1st.Flr.SF+ X2nd.Flr.SF)  %>%   
  filter(!is.na(Bsmt.Full.Bath)) %>%
  filter(!is.na(Bsmt.Half.Bath)) %>%
  mutate(Baths = Bsmt.Full.Bath + 0.5*Bsmt.Half.Bath + Full.Bath + 0.5*Half.Bath) %>%
  mutate(Fireplace.YN = 1*(Fireplaces>0)) %>%
  mutate(Fireplace.Qu =  factor(Fireplace.Qu, levels = levels(addNA(Fireplace.Qu)), labels = c(levels(Fireplace.Qu), "None"), exclude = NULL)) %>%
  mutate(Fireplace.Qu = relevel(Fireplace.Qu, ref="None")) %>%
  mutate(Garage.YN = 1*(!is.na(Garage.Cond))) %>%
  mutate(Garage.Type =  factor(Garage.Type, levels = levels(addNA(Garage.Type)), labels = c(levels(Garage.Type), "None"), exclude = NULL)) %>%
  mutate(Garage.Type = relevel(Garage.Type, ref="None")) %>%
  mutate(Garage.Finish =  factor(Garage.Finish, levels = levels(addNA(Garage.Finish)), labels = c(levels(Garage.Finish), "None"), exclude = NULL)) %>%
  mutate(Garage.Finish = relevel(Garage.Finish, ref="None")) %>%
  mutate(Garage.Qual =  factor(Garage.Qual, levels = levels(addNA(Garage.Qual)), labels = c(levels(Garage.Qual), "None"), exclude = NULL)) %>%
  mutate(Garage.Qual = relevel(Garage.Qual, ref="None")) %>%
  mutate(Garage.Cond =  factor(Garage.Cond, levels = levels(addNA(Garage.Cond)), labels = c(levels(Garage.Cond), "None"), exclude = NULL)) %>%
  mutate(Garage.Cond = relevel(Garage.Cond, ref="None")) %>%
  mutate(Porch.Area = Wood.Deck.SF+ Open.Porch.SF+Enclosed.Porch+X3Ssn.Porch + Screen.Porch) %>%
  mutate(Pool.YN = 1*(Pool.Area>0)) %>%
  mutate(Pool.QC =  factor(Pool.QC, levels = levels(addNA(Pool.QC)), labels = c(levels(Pool.QC), "None"), exclude = NULL)) %>%
  mutate(Pool.QC = relevel(Pool.QC, ref="None")) %>%
  mutate(Fence =  factor(Fence, levels = levels(addNA(Fence)), labels = c(levels(Fence), "None"), exclude = NULL)) %>%
  mutate(Misc.Feature =  factor(Misc.Feature, levels = levels(addNA(Misc.Feature)), labels = c(levels(Misc.Feature), "None"), exclude = NULL)) %>%
  mutate(Mo.Sold = as.factor(Mo.Sold)) %>%
  mutate(Yr.Sold = as.factor(Yr.Sold)) %>%
  dplyr::select(-Garage.Yr.Blt) %>%
  mutate(Condition.1 = as.character(Condition.1)) %>%
  mutate(Kitchen.Qual=plyr::mapvalues(Kitchen.Qual, from = c("Po", "Fa", "TA","Gd", "Ex" ), to = c("1", "2", "3", "4", "5"))) %>%
  mutate(Kitchen.Qual = as.numeric(as.character(Kitchen.Qual))) %>%
  mutate(Heating.QC=plyr::mapvalues(Heating.QC, from = c("Po", "Fa", "TA","Gd", "Ex" ), to = c("1", "2", "3", "4", "5"))) %>%
  mutate(Heating.QC = as.numeric(as.character(Heating.QC))) %>%
  mutate(Bsmt.Qual = droplevels(Bsmt.Qual)) %>%
  mutate(Functional = droplevels(Functional)) %>%
  mutate(Roof.Matl = droplevels(Roof.Matl))



# Simplify Condition 1 (Park, Rail, Normal)
ind_rail<-which(data$Condition.1=="RRNn" | data$Condition.1=="RRAn" | data$Condition.1=="RRNe" | data$Condition.1=="RRAe")
ind_park<-which(data$Condition.1=="PosN" | data$Condition.1=="PosA")
data$Condition.1[ind_rail]<-"Rail"
data$Condition.1[ind_park]<-"Park"
data = data %>% 
  mutate(Condition.1 = factor(Condition.1)) %>%
  mutate(Condition.1 = relevel(Condition.1, ref="Norm")) 

# Eliminate the one entry in 'Exposure' that had been left completely empty
data_train<-data
data_train$Bsmt.Exposure[which(data_train$Bsmt.Exposure=="")]<-"None"
data_train$Bsmt.Exposure<-droplevels(data_train$Bsmt.Exposure)

data_train$Pool.Area<-data_train$Pool.Area+1
data_train$Total.Bsmt.SF<-data_train$Total.Bsmt.SF+1
```

The mantra in real estate appears to be "location, location, location." This begs for a simple visualization of house price distribution by neighborhood.

```{r neighborhood_range, echo=FALSE}
library(ggplot2)
library(treemap)

#Data Vizes
neighborhood.price.range = ggplot(data_train, aes(x = reorder(Neighborhood, desc(price), median), y = price, fill=price)) + geom_boxplot(colour="darkslateblue",fill=terrain.colors(28, alpha=1)) + theme_light() + theme(axis.text.x=element_text(angle=45,hjust=1,vjust=1)) + labs(x = "Neighborhood", y="Price [USD]", title="Boxplot of Price by Neighborhood") + theme(plot.title = element_text(hjust=0.5))

plot(neighborhood.price.range)
```

An interesting observation is that there is a wider dispersion among the more affluent neighborhoods based on relative inter-quartile ranges, whereas the neighborhoods with cheaper housing tend to be more concentrated around their medians. This conforms to a note given to us in the variable guide, which tells us that the assumption of homoskedasticity appears to be violated, as variation increases with sale price. This is an important observation, as it indicates that we would be wise to transform the response variable (which we have, in the form of a log transformation).

With respect to location, we note that Ames is a college town built around Iowa State. With almost 16,000 employees, the unviersity is by far the biggest employer in town. Based on the available neighborhood map, we traced out the distance from the university to the different neighborhoods. This is visualized in the following graph. 


We see that TBD


Our third plot shows the relationship between a house’s price and its overall quality, rated on a scale from 1 (Very Poor) to 10 (Very Excellent).


```{r clusters, echo=FALSE}
library(ggplot2)
ptrain_data = data_train
ptrain_data = ptrain_data%>%mutate(Overall.Qual.level1 = ifelse(Overall.Qual %in% c(1,2,3,4), "low", Overall.Qual))%>%
 mutate(Overall.Qual.level2 = ifelse(Overall.Qual.level1 %in% c("5","6","7"),"med",Overall.Qual.level1))%>%
 mutate(Overall.Qual.level3 = ifelse(Overall.Qual.level2 %in% c("8","9","10"),"high",Overall.Qual.level2))
#sum(ptrain_data$Overall.Qual.level3 %in% c("low","med","high"))

f = qplot(x = TotalSq, y = price, data = ptrain_data, color = factor(Overall.Qual.level3))
f + labs(color='Overall Quality') + xlab("Total Square Feet") + ylab("Price")+ ggtitle("Sale Price versus Total Square Feet by Overall Quality")

```


The red clusters represent the counts of the number of observations that have a certain price and quality, with fatter clusters representing more observations having that price and quality. The gold smooth is a spline fitted with GAM (not by us, but by the geom_smooth function) that runs through the medians of the quality ratings, accompanied by a confidence band in transparent gray. It is interesting to note that most observations fall into the $100,000-300,000 range, as well as the 5 to 8 quality range, which we can observe due to the fatter clusters in these areas. We can also see that the confidence bands through this region are understandably at their most narrow, whereas the bands widen at the Very Poor end of the range, due to there being very few observations at that end. Finally, it is easy to see that there is a non-linear trend in price that occurs around the 7.5 mark, where the slope of the curve steepens. While it is possible that the best fit could be achieved by a spline, logging the price variable seemed to do the trick in correcting for this non-linearity. We did not log Overall Quality, due to its being an integer.


## 2. Development and assessment of an initial model from Part I (10 points)


During the development of our project, we used random forests and boosting not only in our quest to find the best performing model, but also as a useful tool for exploratory data analysis. A particularly useful feature in this regard was the relative importance plot, which revealed that Neighborhood, Lot Area, and Overall Quality were the most important variables. In response to this, we chose to create plots highlighting the relationships between Price and these three important features.


* Initial model: must include a summary table and an explanation/discussion for variable selection.  Interpretation of coefficients desirable for full points.

* Model selection: must include a discussion

* Residual: must include a residual plot and a discussion

* RMSE: must include an RMSE and an explanation  (other criteria desirable)

* Model testing: must include an explanation



## 3. Development of the final model (20 points)

We tried a range of different approaches for a more complex model. Among others, we evaluated tree models, bagging, boosting and random forests, as well as Lasso and Ridge. Based on RMSE however, none of these options was able to outperform a linear model with interaction terms.

In consequence, the final model contains 22 different variables, two 2-way interactions and one 3-way interaction. 

The 10 most imortant varaiables are summarized in the following table

```{r}

```



* Final model: must include a summary table

* Variables: must include an explanation

* Variable selection/shrinkage: must use appropriate method and include an explanation



4. Assessment of the final model (25 points)

* Residual: must include a residual plot and a discussion

* RMSE: must include an RMSE and an explanation  (other criteria desirable)

* Model evaluation: must include an evaluation discussion

* Model testing : must include a discussion

* Model result: must include a selection of the top 10 undervalued and overvalued houses



5. Conclusion (10 points): must include a summary of results and a discussion of things learned





